---
title: "Vignette SICPLN"
author: "Guy Darcy Remesha et Paul-Marie Grollemund"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to SICPLN method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#  SICPLN Method
## Introduction
The SICPLN (Poisson-Lognormal with Smooth Information Criterion) model is designed to model count data, in particular from studies of microbial communities in various environments. This model is used when we want to understand how covariates influence complex biological or ecological phenomena, such as microbial composition in ecosystems linked to milk production.

What sets the SICPLN model apart is its ability to overcome the computational limitations of conventional Poisson-Lognormal models, particularly when the number of covariates is large. By incorporating the Smooth Information Criterion (SIC), this method simplifies variable selection by regularizing model parameters. This leads to a selection of variables that matter most in predictions.

This vignette illustrates the use of the SICPLN model on a dataset, Hspider, and shows the steps involved in selecting variables and predicting the abundance of microbial species in relation to their environments.

# spider study
## data import
```{r ,eval=TRUE,include = TRUE}

suppressPackageStartupMessages(library(VGAM))
data("hspider")
spider <- hspider

```

##Appel du modèle 
```{r ,echo=FALSE,eval=TRUE,include = TRUE}
source("https://raw.githubusercontent.com/pmgrollemund/SICPLN/main/code/SICPLN_Pack.R")
output_dir <- getwd()
```


# Pretreatement 
## Separation of Abundances and Covariates
The PLN model requires two types of data:

1. Response variables: This is a table of count data (matrix), where each row represents an observation and each column represents
a species, product type, etc.

2. Explanatory variables (covariates): These are the factors (continuous or categorical) you want to use to explain variations in the count data.
    
```{r ,echo=FALSE,eval=TRUE,include = TRUE}

# Séparation des données d'abondance et des covariables
Abundance <- spider[,-c(1:6)]
Covariate <- spider[,c(1:6)]

# Parmétrer avec le format de PLN
prep_spider <-  PLNmodels::prepare_data(Abundance,Covariate)

```

## Verification of multicollinearity between covariates
Before building a regression model to analyze abundances, it is important to check that the environmental variables do not exhibit excessive multicollinearity.
```{r ,echo=FALSE,eval=TRUE,include = TRUE}
# calcul des VIF avec variable Reflux
Res_vif_A_Ref <- calculate_VIF(Covariate[,-3], 5)
cat("\n\n\t Variable selection using VIF :\n")
print(Res_vif_A_Ref$selected_variables)
```

### Variables retained after VIF study
```{r ,echo=FALSE,eval=TRUE,include = TRUE}

# Variables retenues après étude des VIF
retain_Covarr <- prep_spider[, Res_vif_A_Ref$selected_variables]
rownames(retain_Covarr) <- rownames(prep_spider$Abundance)
```


```{r ,eval=TRUE,include = TRUE}
data_tmp <- list(Abundance = prep_spider$Abundance, Covariate = retain_Covarr)
data_to_PLN <- prepare_data(data_tmp$Abundance, data_tmp$Covariate)
data_to_PLN <- data_to_PLN[, -ncol(data_to_PLN)]

# Spécifiez les paramètres de contrôle
control <- PLN_param(backend = "nlopt", covariance = "full")
```



## Adjust the PLN model
Once the data is ready, you can adjust the PLN model using the PLN() function. Here's a simple example of how to adjust the model.
```{r ,eval=TRUE,include = TRUE}
# Ajustement du modèle PLN (Poisson-Lognormal) avec la surface en tant qu'offset
cat("\n\n\t launch of the PLN model .\n")
cat("        ==========================\n")

spider_pln <- PLNmodels::PLN(Abundance ~ . , control = control, data = data_to_PLN) 
```




```{r ,eval=TRUE,include = TRUE}
print(spider_pln$model_par$B)
```


## Application du modèle SICPLN
###How to apply the SICPLN method
The SICPLN model is currently available as raw code on GitHub and has not yet been integrated into an R library. However, it works in a similar way to the PLN model, with a few differences in terms of calculation. 

Firstly, the explanatory variables (covariates) and response variables (abundance) are defined from the pre-processed data. The model is fitted using the SICPLN function, which specifies the variable to be used as an offset. 

The default parameters of the PLN model serve as a starting point, but specific control parameters, such as the maximum number of iterations and convergence criteria, can be adjusted. This makes it possible to optimize variable selection by minimizing an information criterion, thus improving the relevance of the results.
```{r ,eval=TRUE,include = TRUE}
# Ajustement pour la selection de variable
cat("\n\n\t launch of the SIC-PLN Model. \n")
cat("        ============================ \n")

formule <- Abundance ~ .
data <- data_to_PLN
spider_sicpln <- SICPLN(formule , data = data_to_PLN, control = control)

```


###  Interpretation of results
After fitting the model, we obtain the most relevant variables influencing microbial abundances.

The model has selected the most important covariates while regularizing the coefficients of irrelevant variables. This makes it possible to manage multicollinearity and improve prediction accuracy.
```{r ,eval=TRUE,include = TRUE}
spider_sicpln$res_fisher$model_par$B
```

# Graphical Results
## B_matrix 
This section shows how to obtain the main graphical results (matrix_B) derived from the PLN and SICPLN methods.

```{r ,eval=TRUE,include = TRUE}

# Graphiques
plot_coef_pln <- coef_genus_barre(spider_pln$model_par$B[-1,], 
                                  axis_names = c("Spider", "Variables", "Coefficients value", "PLN"), data)
plot_coef_pln
plot_coef_sicpln <- coef_genus_barre(spider_sicpln$res_fisher$model_par$B[-1,],
                                     axis_names = c("Spider species", "Variables", "Coefficients value", "SICPLN"), data)
plot_coef_sicpln
```

```{r}
# Save graphics in the specified output folder
ggsave(file.path(output_dir, "plot_pln.pdf"), plot_coef_pln, width = 10, height = 6, units = "in")
ggsave(file.path(output_dir, "plot_sicpln.pdf"), plot_coef_sicpln, width = 10, height = 6, units = "in")
```


## precision matrix
The main purpose of the precision matrix(sigma) is to capture conditional independence between spider species.

The precision matrix presented here for the SICPLN model illustrates the direct conditional relationships between the abundances of different hunter spider species.

```{r ,eval=TRUE,include = TRUE}
# image de la matrice sigma (precision matrix)
tmp <- solve(spider_sicpln$res_fisher$model_par$Sigma)
norm <- sqrt(diag(tmp)) %*% t(sqrt(diag(tmp)))
cor_mat <- tmp / norm
precision_mat_sicpln <- plot_precision_matrix(cor_mat, axis_names = c("", "", "", "SICPLN_Spider precision matrix of Sigma"), limit = NULL)
precision_mat_sicpln
```

## Regression between Abundance and Covariate for significant variables
This graph shows how the abundance of a microbial species varies as a function of an environmental variable, with a trend line to help visualize this relationship.

This function selects variables for which the estimated coefficients (in the B_hat matrix) lie between ]-1, 1[.

For each combination of environmental variable and microbial species (where the coefficient is non-zero), it generates a scatter plot with a fitted linear regression curve.
```{r ,eval=TRUE,include = TRUE}
# # Relation entre variable environnementale et Espèce
plotcovar_Abund_sic <- plot_abundance_vs_environment(spider_sicpln$res_fisher$model_par$B, data, output_dir)

```

# Session informations
```{r session,echo=FALSE,message=FALSE, warning=FALSE}
  sessionInfo()
```
